Variations on a Bayesian Theme: Comparing Bayesian Models of Referential Reasoning

### 2014-03-14 02_RSA.tex
thanks, references, typos

Now let us turn to play as the listener. If we hear ``Square'' or
``Blue'', we easily know the intended referent as there is no
ambiguity, but what if we hear ``Circle'' (or ``Green'')? There are
two circles in the context that we need to choose from. On the one
hand, the blue circle, having the unique color in the context, seems
to be perceptually dominant and thus easily captures our attention. On
the other hand, from the previous analysis we know that if the blue
circle were the intended referent, the speaker could have chosen
``Blue'' which is not ambiguous and thus more informative. Hence the
listener needs to balance two sources of information, i.e., the
(presumably subconscious) perceptual salience of different objects and
the rational expectation of the likelihood of the speaker making the
utterance for each object. The latter line of reasoning is crucial to
the classic Gricean account of scalar implicatures, and the major
challenges in terms of formal modeling are how to quantify notions
such as informativeness and how different pieces of information should
be integrated.

@@

The speaker acts sub-optimally by choosing the utterance that soft-maximizes her expected utility, which is defined as the informativeness of the utterance subtracted by its cost:
\begin{equation} \label{Bayesian_sender_softmax}
\sigma(m \mid t) \propto \exp(\lambda_\mathrm{S}\cdot \mbox{U}(m,t))=\exp(\lambda_\mathrm{S} \cdot (\mbox{Info}(m,t)-\mbox{Cost}(m))),
\end{equation}
where $\lambda_\mathrm{S}$ is a parameter measuring the speaker's
degree of rationality, i.e., to what extent the speaker sticks to the
strict optimum\footnote{Mathematically, it can be shown that when $\lambda_\mathrm{S}=0$, $\sigma(m \mid t)$ is uniformly distributed over all messages, meaning that the speaker just randomly selects one of the messages without any optimality consideration. 
On the other hand, when $\lambda_\mathrm{S}\rightarrow \infty$, 
$\sigma(m \mid t)$ is non-zero only when $m$ yields the maximal utility, 
meaning that the speaker only selects the strictly optimal option(s).
This rule is widely used in the AI literature (e.g., \cite{Sutton1998RL}) and 
is also known as the \emph{exponential rule}.}.
The cost term is used to encode preference in
different utterances, be it about the utterances' lengths or syntactic
categories.

From (\ref{Bayesian_literal})-(\ref{Bayesian_sender_softmax}) we obtain the speaker's production rule\footnote{Note that we can further simplify (\ref{Bayesian_speaker}) as follows: $\exp(\lambda_\mathrm{S} \cdot (\log \mathcal{U}(t\mid \intp{m})-\mbox{Cost}(m)))
=\exp(\lambda_\mathrm{S} \log \mathcal{U}(t\mid \intp{m}) \cdot \exp(-\lambda_\mathrm{S}\cdot \mbox{Cost}(m))
=(\mathcal{U}(t\mid \intp{m}))^{\lambda_\mathrm{S}}\cdot \exp(-\lambda_\mathrm{S}\cdot \mbox{Cost}(m))$. Incidentally, if we do not take costs into account, i.e., let $\mbox{Cost}(m)\equiv 0$, then (\ref{Bayesian_speaker}) reduces to $\sigma(m \mid t) \propto (\mathcal{U}(t\mid \intp{m}))^{\lambda_\mathrm{S}}$, which is effectively a \emph{power rule} that soft-maximizes $\mathcal{U}(t\mid \intp{m})$. In this respect 
(\ref{Bayesian_speaker}) can be seen as an action-oriented speaker model (the exact definition of which will be given in the next section) equipped with 
a different kind of soft-max rule. However, we note that this is a mere technical coincidence, and as a first step, throughout this paper we only consider the exponential 
rule as the soft-max rule. \cite{Worthy2008ratio} provides a detailed discussion and empirical analyses on the differences between the exponential and power rules.}:

>>>>

Now let us turn to play as the listener. If we hear ``Square'' or
``Blue'', we easily know the intended referent as there is no
ambiguity, but what if we hear ``Circle'' (or ``Green'')? There are
two circles in the context that we need to choose from. On the one
hand, the blue circle, having the unique color in the context, seems
to be perceptually dominant and thus easily captures our attention. On
the other hand, from the previous analysis we know that if the blue
circle were the intended referent, the speaker could have chosen
``Blue'' which is not ambiguous and thus more informative. Hence the
listener needs to balance two sources of information, i.e., the
(presumably subconscious) perceptual salience of different objects and
the rational expectation of the likelihood of the speaker making the
utterance for each object. The latter line of reasoning is crucial to
the classic Gricean account of scalar implicatures, and the major
challenges in terms of formal modeling are how to quantify notions
such as informativeness and how different pieces of information should
be integrated.\todo{discuss Frank and Goodman's experiments and
  results (in as far possible) here, before venturing into the model}

@@

The speaker acts sub-optimally by choosing the utterance that soft-maximizes her expected utility, which is defined as the informativeness of the utterance subtracted by its cost:
\begin{equation} \label{Bayesian_sender_softmax}
\sigma(m \mid t) \propto \exp(\lambda_\mathrm{S}\cdot \mbox{U}(m,t))=\exp(\lambda_\mathrm{S} \cdot (\mbox{Info}(m,t)-\mbox{Cost}(m))),
\end{equation}
where $\lambda_\mathrm{S}$ is a parameter measuring the speaker's
degree of rationality, i.e., to what extent the speaker sticks to the
strict optimum\footnote{Mathematically, it can be shown that when $\lambda_\mathrm{S}=0$, $\sigma(m \mid t)$ is uniformly distributed over all messages, meaning that the speaker just randomly selects one of the messages without any optimality consideration. 
On the other hand, when $\lambda_\mathrm{S}\rightarrow \infty$, 
$\sigma(m \mid t)$ is non-zero only when $m$ yields the maximal utility, 
meaning that the speaker only selects the strictly optimal option(s).
This rule is widely used in the AI literature (e.g., \cite{Sutton1998RL}) and 
is also known as \emph{exponential choice rule}.}.
The cost term is used to encode preference in
different utterances, be it about the utterances' lengths or syntactic
categories.

From (\ref{Bayesian_literal})-(\ref{Bayesian_sender_softmax}) we
obtain the speaker's production rule:\footnote{Note that we can further
  simplify (\ref{Bayesian_speaker}) as follows:
  $\exp(\lambda_\mathrm{S} \cdot (\log \mathcal{U}(t\mid
  \intp{m})-\mbox{Cost}(m))) =\exp(\lambda_\mathrm{S} \log
  \mathcal{U}(t\mid \intp{m}) \cdot \exp(-\lambda_\mathrm{S}\cdot
  \mbox{Cost}(m)) =(\mathcal{U}(t\mid
  \intp{m}))^{\lambda_\mathrm{S}}\cdot \exp(-\lambda_\mathrm{S}\cdot
  \mbox{Cost}(m))$. Incidentally, if we do not take costs into
  account, i.e., let $\mbox{Cost}(m) = 0$, then
  (\ref{Bayesian_speaker}) reduces to $\sigma(m \mid t) \propto
  (\mathcal{U}(t\mid \intp{m}))^{\lambda_\mathrm{S}}$, which is
  effectively a \emph{power rule} that soft-maximizes
  $\mathcal{U}(t\mid \intp{m})$. In this respect
  (\ref{Bayesian_speaker}) can be seen as an action-oriented speaker
  model (the exact definition of which will be given in the next
  section) equipped with a different kind of soft-max rule. However,
  we note that this is a mere technical coincidence, and as a first
  step, throughout this paper we only consider the exponential rule as
  the soft-max rule. \cite{Worthy2008ratio} provides a detailed
  discussion and empirical analyses on the differences between the
  exponential and power rules.}

### 2014-03-14 main.tex
minor corrections
several footnotes for reviewers' comments
thanks, references, typos

\title{Variations on a Bayesian Theme: Comparing Bayesian Models of
  Referential Reasoning}
\titlerunning{Variation on a Bayesian Theme}
\author{Ciyang Qing \and Michael Franke\thanks{Thanks to \dots % Judith Degen,
  % Noah Goodman, Daniel Lassiter [?], audience of the workshop, Henk
  % & Hans-Christian for organizing, reviewers \dots 
}}
\institute{Institute for Logic, Language and Computation\\Universtity
  of Amsterdam, The Netherlands}

>>>>

\title{Variations on a Bayesian Theme: Comparing Bayesian Models of
  Referential Reasoning} \titlerunning{Variation on a Bayesian Theme}
\author{Ciyang Qing \and Michael Franke\thanks{We are indebted to
    Judith Degen, Michael C.~Frank, Noah D.~Goodman and Daniel
    Lassiter, two anonymous reviewers, and the audience of the
    ESSLLI workshop ``Bayesian Natural Language Semantics and
    Pragmatics'' for stimulating feedback and discussion. Many thanks
    also to Henk Zeevat and Hans-Christian for organizing mentioned
    workshop, and to Will Frager for help accessing Amazon Mechanical
    Turk. Michael Franke gratefully acknowledges financial support by
    NWO-VENI grant 275-80-004.}}  \institute{Institute for Logic, Language and
  Computation\\Universtity of Amsterdam, The Netherlands}
